A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-04 paint-contour-tree.tex
lower bound changes

\section{Lower Bound by Path Decomposition}
\label{sec:lb}
In this section we provide a lower bound for the time it takes to compute a contour tree in terms the cost of path decompositions.  
We first prove a lower bound for join trees, and then generalize to contour trees.
%\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.

@@

In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
%Observe that given a fixed $(f,\MM)$, 
%our algorithm for computing $\jc$ is deterministic and so the maximum path decomposition $P=P(\jc)$ defined by the initial coloring (\Def{initialColoring}) is fixed, 
%allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
%  
Let $\MM$ be fixed.  For a given height function $f$, we can now define $\cost_{\jc}(f)$.  Specifically, consider running the $\build(f,\MM)$ procedure 
of section \Sec{paint}, which computes the join tree $\jc(f,\MM)$.  This procedure calls $\init$, which will (uniquely) define some 
initial coloring of $\jc$ (\Def{initialColoring}).  Let $P(f)=P(\jc(f,\MM))$ be the maximum path decomposition of the join tree according to this coloring. 
Then $\cost_{\jc}(f) = \sum_{p\in P(f)} |p|\log|p|$

@@

\begin{lemma}
\label{lem:pathDecomp}
Let $\mathcal{P} = P(f)$ be the maximum path decomposition of some height function $f$ defined over a complex $\MM$. 
Then there is a family of functions, $\bF$, all defined over a common triangulation in plane, 
such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $\bF$ define the same join tree.
\end{lemma}
\begin{proof}
 We describe a simple triangulation, with heights specified for all vertices except for the saddles.  The main idea is that 
 this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
 whose size is that described in the lemma statement.

 First, we describe the basic building block of these terrains, which roughly speaking corresponds to a fixed path $p\in P$.  
 Specifically, we say a \emph{diamond} of size $m$ is the following terrain.  To start, construct a slightly tilted cycle 
 of length $m$.  Specifically have two antipodal points, at heights $1$ and $0$, which we will call the anchor and trap of the diamond, respectively.
 The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
 Next, create an apex vertex at some appropriately large height, and add an edge between it and each vertex in the cycle.  At this point 
 we have built an upside down cone with $m$ triangular faces (see \Fig{diamond}).

 Now we describe how to attach two diamonds.  In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
 Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some unspecified height $h$. The remained of the base of the child cone is then 
 extended down (at a slight angle) until it hits the face of the parent.

 At this point we can construct the entire terrain, by repeatedly gluing together diamonds.  
 Specifically, for each path $p_i\in P$ we create a diamond of size $|p_i|+1$.
 The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other diamonds.
 (Note that diamonds have size $|p_i|+1$ since $|p_i|-1$ faces represent the joins of $p_i$, the apex represents the leaf, 
 and we need two empty faces next to the anchor.)
 Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$ (see \Sec{shrubs}).  
 Specially, for two paths $p,q\in P$ where 
 $p$ is the parent of $q$ in $\pathTree$, we glue $q$ onto a face of the diamond for $p$ as described above.
 (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).

@@

 Now there is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
 Let the set of height functions for these instances be denote $\bF$.
 In particular, there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
 functions in this set.  Also, each one of these functions will result in a different join tree.  Specifically, a permutation of the heights on a given diamond 
 corresponds to a permutation of the vertices along a path in $P$, as by construction all saddles on a given diamond will appear in sorted order in the join tree.
\end{proof}

In the following we say to path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
if there is an graph isomorphism, $f$, between their respective shrubs, $S_1$ and $S_2$, which has the property that for any node $v\in S_1$ corresponding to a path of 
size $|v|$, $f(v)$ also corresponds to a path of size $|v|$. Note that for equivalent decompositions, $\sum_{p\in P_1} |p|\log|p| = \sum_{p\in P_2} |p|\log|p|$.

\begin{lemma}
\label{lem:cost}
Let $\mathcal{P}$ and $\bF$ be as defined above.  Then for all $f\in \bF$, $\cost_{\jc}(f) = \sum_{p\in P(f)} |p|\log|p| = \sum_{p\in \mathcal{P}} |p|\log |p|$
\end{lemma}
\begin{proof}
First observe that by construction, regardless of the choice of $f\in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree which is equivalent to $\mathcal{P}$.
Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
We now show that regardless of how paint spilling is done, this will be the maximum path decomposition of $T$ (implying the lemma statement).

 So consider any order of paint spilling over the surface defined by $f$.
 Consider any join on a face of some diamond.  This join is the anchor point of some connecting child diamond.  
 The claim is that the two colors of this join (according to \Def{initialColoring})
 are the colors of the apexes of the child and parent diamond.  
 This easily follows since this join has two up-stars, each of which has exactly one edge.  
 Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
 The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
 so the second color is that of the maxima of the child diamond.

 So let $c$ be the color of some maxima, $m$.  The above proven property of the coloring implies that in the join tree, the vertices for 
 which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
 the vertex to which the root of this path connects to (i.e. the anchor of the diamond of $m$).
 Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
 To see this, consider a vertex $v\in T$, and let $p$ be the path in $\overline{P}(T)$ which contain $v$. 
 Now $v$ has two children (if it is a join).  Let $c_1$ be the child which is in $p$, and $c_2$ the child which is the root of some adjacent path $p'$.  
 In order for $\overline{P}(T)$ to be the maximum path decomposition, we must show $|H_{c_1}|\geq |H_{c_2}|$.
 %(Note, to ensure a unique maximum path decomposition, one can break ties consistently by assigning some arbitrary but fixed ordering to the vertices.)
 %Now $v$ gets assigned two colors, one 
 %of which we know is the color assigned to all vertices in $p$ and the other is the one assigned to all the vertices of $p'$. 
 Observe that none of the colors in the subtree of $c_2$ appear below $v$.  Hence the heap at $c_2$ consists only of $v$.  
 On the other hand the heap of $c_1$ contains $v$, any other vertex of $p$ below $v$, and the vertex to which the root of $p$ is adjacent to, 
 and hence $|H_{c_1}|\geq |H_{c_2}|$.
\end{proof}

>>>>

\section{Lower Bound by Path Decomposition}
\label{sec:lb}
We first prove a lower bound for join trees, and then generalize to contour trees.
%\Ben{Hopefully the following sentence clear up confusion between $\MM$ and $(f,\MM)$}
%Note earlier we usually used $\MM$ to denote the lift of a simplicial complex according to some height function $f$, 
%however in this section we make a distinction between the complex $\MM$ and the height function $f$.  
%In particular all previous definitions or subroutines which took $\MM$ as input now take $(f,\MM)$ as input.

@@

%In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  
%Observe that given a fixed $(f,\MM)$, 
%our algorithm for computing $\jc$ is deterministic and so the maximum path decomposition $P=P(\jc)$ defined by the initial coloring (\Def{initialColoring}) is fixed, 
%allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.
%  
We focus on terrains, so $d=2$.
Consider any valid path decomposition $P$ of a valid join tree (actually, for any rooted binary tree,
there exist a terrain for this is the contour tree).
When we say ``compute the join tree", we require the join tree to
be labeled with the corresponding vertices of the terrain.

%
%For given $\MM$, consider running the algorithm of \Sec{also} to compute $\jc(\MM)$.
%Let $P(\jc(\MM))$ be the maximum path decomposition of the join tree according to the initial coloring
%of $\init(\MM)$. Define $\cost_{\jc}(f) = \sum_{p\in P(\jc(\MM))} |p|\log|p|$

@@

\begin{lemma}
\label{lem:pathDecomp}
Fix a valid path decomposition $P$. There is a family of terrains, $\bF_P$, all with the same triangulation, 
such that $|\bF_P| = \Pi_{p_i\in P} (|p_i|-1)!$, and no two terrains in $\bF_P$ define the same join tree.
\end{lemma}

\begin{proof}
% We describe a simple triangulation, with heights specified for all vertices except for the saddles.

%The main idea is that 
% this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
% whose size is that described in the lemma statement. 
 We describe the basic building block of these terrains, which corresponds to a fixed path $p\in P$.  
 Informally, a \emph{diamond} is an upside down cone with $m$ triangular faces (see \Fig{diamond}). 
 Construct a slightly tilted cycle 
 of length $m$ with the two antipodal points at heights $1$ and $0$. These are called the anchor and trap of the diamond, respectively.
 The remaining $m-2$ vertices are evenly spread around the cycle and heights decrease monotonically when going from the anchor to the trap.
 Next, create an apex vertex at some appropriately large height, and add an edge each vertex in the cycle.

 Now we describe how to attach two different diamonds.  In this process, we glue the base of a scaled down ``child'' diamond on to a triangular cone face of the larger  ``parent'' diamond (see \Fig{diamond}).
 Specifically, the anchor of the child diamond is attached directly to a face of the parent diamond at some height $h$. The remaineder of the base of the child cone is then 
 extended down (at a slight angle) until it hits the face of the parent.

 The full terrain is obtained by repeatedly gluing diamonds. For each path $p_i\in P$, we create a diamond of size $|p_i|+1$.
 The two faces adjacent to the anchor are always empty, and the remaining faces are for gluing on other diamonds.
 (Note that diamonds have size $|p_i|+1$ since $|p_i|-1$ faces represent the joins of $p_i$, the apex represents the leaf, 
 and we need two empty faces next to the anchor.)
 Now we glue together diamonds of different paths in the same way the paths are connected in the shrub $\pathTree$ (see \Sec{shrubs}).  
 Specially, for two paths $p,q\in P$ where 
 $p$ is the parent of $q$ in $\pathTree$, we glue $q$ onto a face of the diamond for $p$ as described above.
 (Naturally for this construction to work, diamonds for a given path will be scaled down relative to the size of the diamond of their parent).
 By varying the heights of the gluing, we get the family of terrains.

@@

 There is a valid instance of this described construction for any relative ordering of the heights of the saddles on a given diamond.  
 In particular, there are $(|p|-1)!$ possible orderings of the heights of the saddles on the diamond for $p$, and hence $\Pi_{p_i\in P} (|p_i|-1)!$ possible 
 terrains we can build.  Each one of these functions will result in a different (labeled) join tree. All saddles on a given diamond will appear in sorted order in the join tree.
 So, any permutation of the heights on a given diamond corresponds to a permutation of the vertices along a path in $P$.
\end{proof}

Two path decompositions $P_1$ and $P_2$ (of potentially different complexes and/or height functions) are equivalent 
if: there is a 1-1 correspondence between the sizes of the constituent paths, and the shrubs are isomorphic.
%
%
%if their respective shrubs, $S_1$ and $S_2$, which has the property that for any node $v\in S_1$ corresponding to a path of 
%size $|v|$, $f(v)$ also corresponds to a path of size $|v|$. Note that for equivalent decompositions, $\sum_{p\in P_1} |p|\log|p| = \sum_{p\in P_2} |p|\log|p|$.

\begin{lemma}
\label{lem:cost}
For all $\MM \in \bF_P$, the total number of heap operations performed by $\build(\MM)$ is $\Theta(\sum_{p\in P} |p|\log|p|)$.
\end{lemma}
\begin{proof}
%For any choice of $\MM \in \bF$, there is a path decomposition $\overline{P}(T)$ of the join tree equivalent to $\mathcal{P}$.
%Specifically, this is the decomposition where each path $p\in \overline{P}(T)$ consists exactly of the set of critical points (except for its anchor) from a given diamond. 
The primary ``non-determinism" of the algorithm is the initial painting constructed by $\init(\MM)$. 
We show that regardless of how paint spilling is done, the number of heap operations is bounded as above.

 Consider an arbitrary order of the initial paint spilling over the surface defined by $f$.
 Consider any join on a face of some diamond, which is the anchor point of some connecting child diamond.  
 The join has two up-stars, each of which has exactly one edge. Each edge connects to a maxima
 and must be colored by that maxima.
 Hence, two colors touching this join (according to \Def{initialColoring})
 are the colors of the apexes of the child and parent diamond.  
% 
% Specifically, one of these edges connects directly to the apex of the parent diamond and so the join must get the color of this maxima.
% The other edge connects directly to the apex of the child diamond (this is why we left the faces adjacent to an anchor empty) and 
% so the second color is that of the maxima of the child diamond.

 Take any join $v$, with two children $w_1$ and $w_2$. Suppose $w_1$ and $v$ belong to the same path
 in the decomposition. The key is that any color from a maxima in the subtree at $w_2$ cannot touch
 any ancestor of $v$. This subtree is exactly the contour tree of the child diamond attached at $v$.
 The base of this diamond is completely contained in a face of the parent diamond. So all colors
 from the child ``drain off" to the base of the parent, and do not touch any joins on the parent diamond.

 Hence, $|H_v|$ is at most the size of the path in $P$ containing $v$. By \Lem{runTimeUpper}, the total number of heap operations
 is at most $\sum_v \log |H_v|$, completing the proof.

% 
% So let $c$ be the color of some maxima, $m$.  The above proven property of the coloring implies that in the join tree, the vertices for 
% which $c$ appears are exactly the vertices of the path in  $\overline{P}(T)$ corresponding $m$, as well as 
% the vertex to which the root of this path connects to (i.e. the anchor of the diamond of $m$).
% Now we show this implies $\overline{P}(T)$ is a maximum path decomposition.
% To see this, consider a vertex $v\in T$, and let $p$ be the path in $\overline{P}(T)$ which contain $v$. 
% Now $v$ has two children (if it is a join).  Let $c_1$ be the child which is in $p$, and $c_2$ the child which is the root of some adjacent path $p'$.  
% In order for $\overline{P}(T)$ to be the maximum path decomposition, we must show $|H_{c_1}|\geq |H_{c_2}|$.
% %(Note, to ensure a unique maximum path decomposition, one can break ties consistently by assigning some arbitrary but fixed ordering to the vertices.)
% %Now $v$ gets assigned two colors, one 
% %of which we know is the color assigned to all vertices in $p$ and the other is the one assigned to all the vertices of $p'$. 
% Observe that none of the colors in the subtree of $c_2$ appear below $v$.  Hence the heap at $c_2$ consists only of $v$.  
% On the other hand the heap of $c_1$ contains $v$, any other vertex of $p$ below $v$, and the vertex to which the root of $p$ is adjacent to, 
% and hence $|H_{c_1}|\geq |H_{c_2}|$.
\end{proof}
